---
title: "Untitled"
author: "Jones Kurian"
date: "09/09/2022"
output: pdf_document
---

```{r}
dat=read.table('assignment2_prob1.txt',header = T)
head(dat)
dat$duration <- factor(dat$duration, levels=c("0-4","5-9","10-14","15-19","20-24","25-29"))
dat$residence <- factor(dat$residence, levels=c("Suva", "urban", "rural"))
dat$education <- factor(dat$education, levels=c("none", "lower", "upper", "sec+"))
```
"because the number of children that can be birthed by a number of women does not have strict
limits, a binomial distribution does not work. I will try a poisson instead."

```{r}
with(data = dat,interaction.plot(duration,education,log(nChildren)))
with(data = dat,interaction.plot(duration,residence,log(nChildren)))
with(data = dat,interaction.plot(education,residence,log(nChildren)))
```
'clear evidence of interaction. I took the log of children because that is the link fnuction I am using'
'the other 2 pairs do not show signs of interaction'

```{r}
pmod=glm(nChildren~nMother+duration*education+residence,family=poisson,data=dat)
summary(pmod)
#The model still lacks adequacy, will check the pearson chi stat'
sum(residuals(pmod,'pearson')^2)/43
```
'the dataset is wildly overdispersed. Instead of transforming this model, I will choose a gamma dist,
because that has more paramatres and is similar enough to a poisson dist. The gamma distribution
is also advantageuos because unlike the poisson it can take continuous values instead of only discrete, so
i will make a model with fertility as the dependant variable.'
```{r}
dat$fertility=dat$nChildren/dat$nMother
gmod=glm(fertility~duration+residence+education,family = Gamma,data = dat)
summary(gmod)
step(gmod)
```

```{r}
with(data = dat,interaction.plot(duration,residence,predict(gmod)))
with(data = dat,interaction.plot(education,residence,predict(gmod)))
with(data = dat,interaction.plot(education,duration,predict(gmod)))
```
'no signs of interaction'
'I am now checking the distribution of the residuals to see if the model is good and how the 
residuals behave'
```{r}
plot(gmod)
```
'broadly speaking the errors seem well behaved the pearsons residuals look normally distributed.
Points 17 and 56 may be outliers'
```{r}
plot(dat$fertility~predict(gmod,dat,'response'))
abline(a=0,b=1,col='red')
```
'the plot of predicted vs fitted should match the line y=x. This does seem to be 
the case. The data seems to be clustered around the fit. The model seems pretty 
good. I will remove the suspected outliers (17&56) and see what happens'
```{r}
gmod2=glm(fertility~duration+residence+education,family = Gamma,data = dat[c(-17,-56),])
summary(gmod2)
plot(gmod2)
```
'i think the plots for gmod2 are a little worse. Looking at the q-q plot the 
residuals appear less normal but the residuals do appear more heteroskedastic. 
I will only remove 56 and see what happens'
```{r}
gmod3=glm(fertility~duration+residence+education,family = Gamma,data = dat[c(-56),])
plot(gmod3)
1-sum((dat$fertility-predict(gmod2,dat,'response'))^2)/sum((dat$fertility-mean(dat$fertility))^2)
1-sum((dat$fertility-predict(gmod3,dat,'response'))^2)/sum((dat$fertility-mean(dat$fertility))^2)
1-sum((dat$fertility-predict(gmod,dat,'response'))^2)/sum((dat$fertility-mean(dat$fertility))^2)
```
'gmod3 does appear to be an improvement from gmod2 and gmod1. I will select gmod3. The numbers above are similar to an R^2 statistic'
```{r}
plot(dat$fertility~predict(gmod3,dat,'response'),pch=16,col=dat$residence)
plot(dat$fertility~predict(gmod3,dat,'response'),pch=16,col=dat$education)
plot(dat$education,1/dat$fertility)
#from both the plots it would seem the outliers are in education and residence This is surprising 
#because the interaction plot seemed parrallel.
gmod4=glm(fertility~education*residence+duration,family=Gamma,data = dat)
summary(gmod4)
```
'basically only sec+ and urban interact all the others are non-significant. This explains 
why I failed to notice the interaction the first time around because except for that pair all the others
are parrallel.'

```{r}
fstat=(deviance(gmod3)-deviance(gmod4))/6/sum(residuals(gmod4,'pearson')^2/53)
pf(fstat,6,53,lower.tail = F)#calculated p value for whether gmod4 is better than gmod3
step(gmod4)
plot(dat$fertility~predict(gmod4,dat,'response'))
abline(a=0,b=1)
plot(dat$fertility~predict(gmod3,dat,'response'))
abline(a=0,b=1)
```
'this is not good, the gmod3 excludes 1 of the outlier points in its calculation of deviance 
but gmod4 includes all points given how close the p value is to 0.05 it is possible that 
this single point will make the difference. Given how close it is and the result of the step
function which says all the variables in gmod4 are worth keeping, I leaning towards gmod4 as the 
best model.'

```{r}
summary(lm(dat$fertility~predict(gmod4,dat,'response')-1))
summary(lm(dat$fertility~predict(gmod3,dat,'response')-1))
plot(gmod4)
1-sum((dat$fertility-predict(gmod3,dat,'response'))^2)/sum((dat$fertility-mean(dat$fertility))^2)
1-sum((dat$fertility-predict(gmod4,dat,'response'))^2)/sum((dat$fertility-mean(dat$fertility))^2)
```
'gmod4 seems much better. I think gmod4 is the best model for representing the relationship between the 
various factors and fertility.'
